Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-14 performance.tex
summary

\section{SUMMARY}

With Bempp-cl we have created a Python platform that achieves high-performance through use of modern just-in-time compilation technologies. Bempp-cl mixes Numba evaluation for less compute intensive linear complexity loops and sparse matrix generation with highly optimized OpenCL kernels for computationally sensitive dense matrix assembly routines. Basing development on Python instead of classical C/C++ makes it very easy to adapt the library and integrate into other applications with Python interfaces. An example is the recent work of integrating Bempp-cl and the Exafmm library to build electrostatic virus-scale simulations \cite{bempp_exafmm}. Strict separation of the computational backend from the interface layer in the library also makes it easy to integrate further backends. We currently support Numba and OpenCL as compute backends. Other technologies could be added easily. OpenCL itself has proved a valuable choice for the purpose of this library. We have all kernels available in CPU and GPU optimized versions and the user can easily move between CPU and GPU based compute devices by a simple parameter change, allowing us to support a wide range of hardware. Here, we have demonstrated Nvidia benchmarks. But it would have made no difference to use AMD or Intel GPUs.

A disadvantage of our approach is that using OpenCL kernels is mixed-language development due to the OpenCL kernels written in C99. Using Numba throughout would be a much more native Python experience. However, while Numba is constantly improving, achieving optimal performance for complex operations is difficult. OpenCL shines here. It makes explicit SIMD operations very easy through dedicated constructs. Moreoever, OpenCL kernels are completely stack based functions, allowing much better compiler optimisations while Numba needs to create every object dynamically, even very small arrays for objects such as coordinate vectors.

Also, one has to consider the type of operations that are required to be accelerated. For the dense assembly of integral operators we have very simple data structures that can easily be passed to compute kernels. More complex data structures with larger mixture of data movement operations and computations (e.g. tree-based algorithms), are much harder to accelerate since the Python layer imposes limits here on the performance.

Overall, with the model of mixed Python/OpenCL/Numba development we have created a flexible and easy to extend platform for integral equation computations and it was an effort that has payed back in terms of simple development without sacrificing performance. Strict separation of compute backends and higher level routines makes it easy for us to integrate other accelerator techniques in the future with little code changes, and to react to new trends in heterogeneous computing.

The current focus of further developments is on letting Bempp-cl take advantage of cluster computing by integrating the \verb@mpi4py@ MPI bindings for Python. We have also made big steps forward for large problems by creating a black-box FMM interface that is currently implemented for Exafmm, which has allowed us to solve problems with 10 million elements on a single workstation. This shows that Python focused development (potentially with some native routines in lower-level languages) is a scalable model and we are aiming to exploit this scalability further as we move from single workstation computations to large cluster problems.

>>>>

\section{SUMMARY}
With Bempp-cl, we have created a Python library that achieves high-performance through use of modern just-in-time compilation technologies. Bempp-cl mixes Numba evaluation for less compute intensive linear complexity loops and sparse matrix generation with highly optimized OpenCL kernels for computationally sensitive dense matrix assembly routines. Basing development in Python instead of classical C/C++ makes it very easy to adapt the library and integrate with other libraries with Python interfaces, such as the recent work integrating Bempp-cl and the Exafmm library to build electrostatic virus-scale simulations \cite{bempp_exafmm}.

Strict separation of the computational backend from the interface layer in the library also makes it easy to integrate further backends, allowing us to remain up to date with future compute models. OpenCL itself has proved a valuable choice for the purpose of this library, as it allows us to run CPU and GPU optimized kernels with very little code overhead, and allows the user to easily move between CPU and GPU based compute devices with a simple parameter change. In this article, we demonstrated Nvidia benchmarks, the same benchmarking code could be used to run on AMD or Intel GPUs (for example).

A disadvantage of our approach is that using OpenCL kernels introduce a second language (C99) to the library. Using Numba throughout would give a much more native Python experience, but, while Numba is constantly improving, it is currently difficult to achieve optimal performance for complex operations. OpenCL really shines here, as it makes explicit SIMD operations very easy through dedicated constructs. Moreoever, OpenCL kernels are completely stack based functions, allowing much better compiler optimisations while Numba needs to create every object dynamically, even very small arrays for objects such as coordinate vectors.

Additionally, one has to consider the type of operations that need to be accelerated. For the dense assembly of integral operators, we have very simple data structures that can easily be passed to compute kernels. More complex data structures with larger mixture of data movement operations and computations (e.g., tree-based algorithms), are much harder to accelerate since the Python layer imposes limits here on the performance.

Overall, with the model of mixed Python/OpenCL/Numba development, we have created a flexible and easy to extend platform for integral equation computations. The initial re-implementation efforts are really paying off, as they allow us to develop new features in a far simpler environment without sacrificing performance. Strict separation of compute backends and higher level routines makes it easy for us to integrate other accelerator techniques in the future with little code changes, and to react to new trends in heterogeneous computing.

The current focus of further developments is on letting Bempp-cl take advantage of cluster computing by integrating the \pyth{mpi4py} MPI bindings for Python. We have also made big steps forward for large problems by creating a black-box FMM interface that currently intenfaces to Exafmm: this has allowed us to solve problems with 10 million elements on a single workstation. We believe that Python-focused development (with some native routines in lower-level languages) is a scalable model and are aiming to exploit this scalability further as we move from single workstation computations to large cluster problems.
